Galois Slicing as Differentiation

### 2025-07-11 main/introduction.tex
WIP.

Existing approaches to program slicing are often tied to particular programming languages or implementations.
In this paper we develop a general categorical approach to program slicing, focusing on a particular technique
called \GPS, where the forward and backward slicing procedures for a particular execution form a Galois
connection. Our main insight is that this approach to slicing can be seen as a generalised form of
differentiable programming, with approximations (slices) of elements playing the role of tangents.

@@

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for \GPS.
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
\item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
\item We also relate \GPS to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}

>>>>

Existing approaches to program slicing are often tied to particular programming languages or implementations.
In this paper we develop a general categorical approach to program slicing, focusing on a particular technique
called \GPS, where the set of slices of a given value form a lattice of approximations and the forward and
backward slicing procedures give rise to Galois connections between these lattices. Our main insight is that
this approach to slicing can be seen as a generalised form of differentiable programming, with approximations
(slices) of elements playing the role of tangents. Our categorical approach should provide a suitable setting
for exploring data provenance for a variety of programming languages, and also allows easy experimentation
with alternative approximation strategies, including quantitative forms of slicing.

@@

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for \GPS.
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
\item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
% \item We also relate \GPS to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}
